Paper 2: survey analysis

### 2024-10-08 Rmd/Paper2.Rmd
Removed reanalysis done for synopsis

---
title:        "Paper 2: survey analysis"
author:       "Niklas Karlsen"
date:         "2024-09-05"
bibliography: "../data/meta_data/references.bib"
csl:          "../scripts/university-of-gothenburg-apa-7th-edition-swedish-legislations.csl" # (source: https://bookdown.org/yihui/rmarkdown-cookbook/bibliography.html)
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true

@@

# cluster-variable
# kilde: https://www.r-bloggers.com/2021/05/clustered-standard-errors-with-r/
library(estimatr)                           # for cluster corrected SE
#library(sandwich)
#library(lmtest)
library(nlme)                               # for ICC - variance HLM
```

@@

## Reanalysis of OLS

A reanalysis of the OLS is done here, due to some methodical reconsiderations after a closer look at the MLM. The dataset is slightly closer to the rawdata (3 NAs were reset to other, as in the original dataset). The order of the variables are also slightly different (TEIs are introduced before covariates). Finally, interaction effects are not included, as they were in paper 2. The statistical tests suggested that the interaction effects were non-significant. Otherwise, the analysis is similar to paper 2.

### Reload data

```{r}
# ref: https://www.statology.org/clear-environment-in-r/ [5.9.2024]
#clear all data frames from environment
rm(list=ls(all=TRUE)[sapply(mget(ls(all=TRUE)), class) == "data.frame"])
```

Then I reload the dataset from file.

```{r load_data_reanalysis}
# Rdata-file contains reduced / prepared datasets
load(file="../data/processed_data/tpack-TRELIS-Q_complete_2023-paper2.Rdata")
```

Then I recenter the data.

And add some missing information.

NMAR are removed.

```{r fix_missing_2, eval=T}
# remove NMAR + Institute9 (19)
data_clean <- data_TRELIS_Q[-c(19,34,44,45,46,95,102),] 
```

#### baseline model (nullmodell)

```{r TK_baseline}
# kovariater; we use scale to standardize when using interaction effects
# sjekk: er det riktig å skalere?
model_TK_baseline <- lm_robust(TK ~ factor(studiested), 
                clusters = factor(studiested), data=data_clean)
summary(model_TK_baseline)
```

```{r TKlevel1_main}
# kovariater; we use scale to standardize when using interaction effects
# sjekk: er det riktig å skalere?
model_TK_level1 <- lm_robust(TK ~ factor(studiested) + factor(Gender) + (factor(Trinn) + factor(semester)), 
                clusters = factor(studiested), data=data_clean)
summary(model_TK_level1)
```

```{r}
# source: https://www.statology.org/how-to-calculate-the-p-value-of-an-f-statistic-in-r/
# source: p. 257 i Howitt & Cramer, 2014
fstat <- model_TK_level1$adj.r.squared / model_TK_baseline$adj.r.squared
df1   <- length(data_clean$TK) - 1
df2   <- df1
pf(fstat, df1, df2, lower.tail = FALSE)
```

The F-statistic indicate that the two models are not significantly different. This means that level1 effects do not contribute significantly to the variance.

#### Baseline

```{r PCKlevel2_main}
# studiested; Institute1 velges som referansekategori siden den har lavest TPACK-verdi av alle
model_PCK_baseline <- lm_robust(
  PCK ~ 
    factor(studiested),                                 # level 2
  clusters = factor(studiested), data = data_clean
)
summary(model_PCK_baseline)
```

```{r PCKlevel1_main}
# kovariater; we use scale to standardize when using interaction effects
# sjekk: er det riktig å skalere?
model_PCK_level1 <- lm_robust(PCK ~ factor(studiested) + factor(Gender) + factor(Trinn) + factor(semester),
  clusters = factor(studiested), data = data_clean
)
summary(model_PCK_level1)
```

```{r}
# source: https://www.statology.org/how-to-calculate-the-p-value-of-an-f-statistic-in-r/
# source: p. 257 i Howitt & Cramer, 2014
fstat <- model_PCK_level1$adj.r.squared / model_PCK_baseline$adj.r.squared
df1   <- length(data_clean$PCK) - 1
df2   <- df1
pf(fstat, df1, df2, lower.tail = FALSE)
```

There were significant differences when including covariates.

#### Baseline

```{r level2_main}
# studiested; Institute1 velges som referansekategori siden den har lavest TPACK-verdi av alle
model_TPACK_baseline <- lm_robust(
  TPACK ~
    factor(studiested),                                 # level 2
  clusters = factor(studiested), data = data_clean
)
summary(model_TPACK_baseline)
```

```{r level1_main}
# kovariater; we use scale to standardize when using interaction effects
# sjekk: er det riktig å skalere?
model_TPACK_level1 <- lm_robust(
  TPACK ~
    factor(studiested) + factor(Gender) + factor(Trinn) + factor(semester),
  clusters = factor(studiested), data = data_clean
)
summary(model_TPACK_level1)
```

```{r}
# source: https://www.statology.org/how-to-calculate-the-p-value-of-an-f-statistic-in-r/
# source: p. 257 i Howitt & Cramer, 2014
fstat <- model_TPACK_level1$adj.r.squared / model_TPACK_baseline$adj.r.squared
df1 <- length(data_clean$TPACK) - 1
df2 <- df1
pf(fstat, df1, df2, lower.tail = FALSE)
```

There is no significant difference when including covariates.

```{r PCK}
# PCK
model_TPACK_PCK <- lm_robust(
  TPACK ~
    factor(Gender) + factor(Trinn) + factor(semester) + # level 1
    factor(studiested) +                                # level 2
    PCK,
  clusters = factor(studiested), data = data_clean
)
summary(model_TPACK_PCK)

# main
fstat <- model_TPACK_PCK$adj.r.squared / model_TPACK_level1$adj.r.squared
pf(fstat, df1, df2, lower.tail = FALSE)

```{r TK}
# TK
model_TPACK_TK <- lm_robust(
  TPACK ~
    factor(Gender) + factor(Trinn) + factor(semester) + # level 1
    factor(studiested) +                                # level 2
    PCK + TK,
  clusters = factor(studiested), data = data_clean
)
summary(model_TPACK_TK)

The results differ somewhat from paper 2 in that some interaction effects are excluded, which was included in the analysis done for paper 2. The main trends are however similar.

mtkcl1rs  <- model_TK_baseline$r.squared
mtkcl1ars <- model_TK_baseline$adj.r.squared

mtkcl2rs  <- model_TK_level1$r.squared
mtkcl2ars <- model_TK_level1$adj.r.squared

rownames(TK_R2) <- c("Teacher education institute", "Gender, year of study, target grade level")

mpcl1rs   <- model_PCK_baseline$r.squared
mpcl1ars  <- model_PCK_baseline$adj.r.squared

mpcl2rs   <- model_PCK_level1$r.squared
mpcl2ars  <- model_PCK_level1$adj.r.squared

rownames(PCK_R2) <- c("Teacher education institute", "Gender, year of study, target grade level")

mtcl1rs  <- model_TPACK_baseline$r.squared
mtcl1ars <- model_TPACK_baseline$adj.r.squared

mtcl2rs  <- model_TPACK_level1$r.squared
mtcl2ars <- model_TPACK_level1$adj.r.squared

rownames(TPACK_R2) <- c("Teacher education institute", "Gender, year of study, target grade level", "PCK", "TK")

Level 1: gender, target grade level  
Level 2: year of study, institute

Burde semester vært en level 2 variabel, siden det er mer en egenskap ved instituttet enn studenten?

Spørsmålet er om forskjeller på semester skyldes endringer på instituttet (level 2) eller endringer i studentmassen (level 1)? Både-og? Samspill på tvers av nivåer!

## Multi-level modeling with cross-level interaction effects

When the data is nested, multi-level modeling can be used. However, when the number of level 2-units is below 10, as it is in our dataset (we have only 5 TEIs), OLS with cluster corrected standard errors can be used. I, however, do a MLM here to compare with the OLS results. MLM is only done for TPACK as dependent variable to illustrate the method.

The r package NLME [@pinheiro_nlme_2023] is used for the analysis.

### Baseline model

First, the baseline model is calculated as a reference and to find the ICC.

```{r}
# baseline model / nullmodell

# REML is used for parameter estimation
model_MLM_baseline <- lme(TPACK~1, data=data_clean, method="REML", na.action="na.omit", random= ~1|studiested)

# ML is used for model fit
model_MLM_baseline_fit <- lme(TPACK~1, data=data_clean, method="ML", na.action="na.omit", random= ~1|studiested)

summary(model_MLM_baseline)
summary(model_MLM_baseline_fit)
```

```{r}
# ICC
nu_00 <- var(ranef(model_MLM_baseline)) # intercept (i.e., TEIs)
nu_e  <- var(resid(model_MLM_baseline)) # residuals (i.e., PSTs)
ICC   <- nu_00 / (nu_00 + nu_e)
ICC
```

30 % av variansen i TPACK ligger på nivå 2. Dette er større enn den anbefalte grensen på 5 %. Vurder om OLS-regresjon med klyngekorrigert standardfeil egner seg bedre enn flernivåanalyse [@christophersen_introduksjon_2018] (p. 109).

```{r}
# effektiv utvalgsstørrelse
n_e <- 147 / (1 + ((147 / 5) - 1) * ICC)
n_e
```

Utvalgsstørrelsen korrigert for avhengighet er lik 15.

### Level 1 model (random intercept)

The level 1 model extends the baseline model by including the covariates.

#### Level 1 variables

```{r}
# REML is used for parameter estimation
model_MLM_level1 <- lme(TPACK~Gender + semester + Trinn, data=data_clean, method="REML", na.action="na.omit", random= ~1|studiested)

# ML is used for model fit
model_MLM_level1_fit <- lme(TPACK~Gender + semester + Trinn, data=data_clean, method="ML", na.action="na.omit", random= ~1|studiested)

summary(model_MLM_level1)
summary(model_MLM_level1_fit)
```

Er level 1-modellen bedre enn nullmodellen?

```{r eval=T}
anova(model_MLM_baseline_fit, model_MLM_level1_fit)
```

#### Level 2 variables

Since Level 2 variables are unique to MLM, and the solution does not converge, level 2 analysis is not done here.

La kovariatene variere mellom studiesteder

```{r eval=F}
# REML is used for parameter estimation
model_level2 <- lme(TPACK~Gender + semester + Trinn, data=data_clean, method="REML", na.action="na.omit", random= ~ semester|studiested)

# ML is used for model fit
model_level2_fit <- lme(TPACK~Gender + semester + Trinn, data=data_clean, method="ML", na.action="na.omit", random= ~ semester|studiested)

summary(model_level2)
summary(model_level2_fit)
```

Er level 2-modellen bedre enn level 1?

```{r eval=F}
anova(model_level2_fit, model_MLM_level1_fit)
```

### PCK

```{r}
# grand mean centering: scale(TK, center = T, scale = F)

data_clean$PCK_cent <- scale(data_clean$PCK, center = T, scale = F)
data_clean$TK_cent  <- scale(data_clean$TK, center = T, scale = F)

# REML is used for parameter estimation
model_MLM_PCK <- lme(TPACK~Gender + semester + Trinn + PCK_cent, data=data_clean, method="REML", na.action="na.omit", random= ~1|studiested)

# ML is used for model fit
model_MLM_PCK_fit <- lme(TPACK~Gender + semester + Trinn + PCK_cent, data=data_clean, method="ML", na.action="na.omit", random= ~1|studiested)

summary(model_MLM_PCK)
summary(model_MLM_PCK_fit)
```

```{r eval=T}
anova(model_MLM_PCK_fit, model_MLM_level1_fit)
```

### TK

```{r}
# REML is used for parameter estimation
model_MLM_TK <- lme(TPACK~Gender + semester + Trinn + PCK_cent + TK_cent, data=data_clean, method="REML", na.action="na.omit", random= ~1|studiested)

# ML is used for model fit
model_MLM_TK_fit <- lme(TPACK~Gender + semester + Trinn + PCK_cent + TK_cent, data=data_clean, method="ML", na.action="na.omit", random= ~1|studiested)

summary(model_MLM_TK)
summary(model_MLM_TK_fit)
```

```{r eval=T}
anova(model_MLM_PCK_fit, model_MLM_TK_fit)
```

## Comparison of OLS and MLM

### Comparison of variance

I follow @christophersen_introduksjon_2018 (p. 115) and use the variance estimates as an indicator of R2.

```{r}
# MLM
# variance baseline
nu_00_baseline <- var(ranef(model_MLM_baseline)) # intercept (i.e., TEIs)
nu_e_baseline  <- var(resid(model_MLM_baseline)) # residuals (i.e., PSTs)

# variance level 1
nu_00_level1   <- var(ranef(model_MLM_level1)) # intercept (i.e., TEIs)
nu_e_level1    <- var(resid(model_MLM_level1)) # residuals (i.e., PSTs)

# variance PCK
nu_00_PCK      <- var(ranef(model_MLM_PCK)) # intercept (i.e., TEIs)
nu_e_PCK       <- var(resid(model_MLM_PCK)) # residuals (i.e., PSTs)

# variance TK
nu_00_TK       <- var(ranef(model_MLM_TK)) # intercept (i.e., TEIs)
nu_e_TK        <- var(resid(model_MLM_TK)) # residuals (i.e., PSTs)

# OLS
mtcl1ars       <- model_TPACK_baseline$adj.r.squared
mtcl2ars       <- model_TPACK_level1$adj.r.squared
mtpars         <- model_TPACK_PCK$adj.r.squared
mttars         <- model_TPACK_TK$adj.r.squared

# table of difference in variance

table_variance <- data.frame(
  c(
    nu_e_baseline,
    (nu_e_baseline - nu_e_level1) / nu_e_baseline,
    ((nu_e_baseline - nu_e_PCK) / nu_e_baseline) - ((nu_e_baseline - nu_e_level1) / nu_e_baseline),
    ((nu_e_baseline - nu_e_TK) / nu_e_baseline) - ((nu_e_baseline - nu_e_PCK) / nu_e_baseline) - ((nu_e_baseline - nu_e_level1) / nu_e_baseline)
  ),
  c(
    nu_00_baseline,
    (nu_00_baseline - nu_00_level1) / nu_00_baseline,
    ((nu_00_baseline - nu_00_PCK) / nu_00_baseline) - ((nu_00_baseline - nu_00_level1) / nu_00_baseline),
    ((nu_00_baseline - nu_00_TK) / nu_00_baseline) - ((nu_00_baseline - nu_00_PCK) / nu_00_baseline) + ((nu_00_baseline - nu_00_level1) / nu_00_baseline)
  ),
  c(mtcl1ars, mtcl2ars - mtcl1ars, mtpars - mtcl2ars, mttars - mtpars)
)

rownames(table_variance) <- c("TEI", "covariates", "PCK", "TK")
colnames(table_variance) <- c("MLM Level 1 % var", "MLM Level 2 % var", "OLS reanalysis diff R2")

pander(table_variance)
```

The variance differ somewhat from the standard deviation (when squared) reported in the summary of the MLMs. Why? Is the OLS a kind of average of the MLM, which is separated into level 1 and level 2?

@@

The reanalysis using OLS gave similar results as for paper 2. Only TEIs explained variance in TK, not the covariates. The covariates did however significantly explain the variance in PCK, but the TEIs did not (or to a small degree, only 3%). The variance in TPACK was significantly explained by TEI, PCK and TK, not the covariates.

The results in paper 2 indicated that the covariates contributed to explain the variance in TPACK. The reanalysis done here, suggest that this may be due to the order in which the variables were added to the model being tested. If I follow the order used in the MLM, the reanalysis suggests that the covariates do not contribute to TPACK.

The MLM gave supporting results. Including the covariates did not improve the model, whereas including PCK and TK did. The variance at level 2 (between TEIs) was due to TK, not PCK or covariates, that is, there were significant differences in TK between TEIs. At level 1 (between PSTs) 6% of the variance were due to the covariates, 17 % due to PCK and 55 % due to TK.

The OLS reanalysis suggested that 30 % of the variance in TPACK was due to TEIs, 11 % due to PCK and 32 % due to TPACK. The covariates did not contribute to the variance in TPACK. Even though the percentages differ between the OLS and the MLM, they show the same trends. The different percentages may be due to how R2 (or difference in variance) is calculated differently in OLS and MLM.

>>>>

---
title:        "Paper 2: survey analysis"
author:       "Niklas Karlsen, OsloMet, Norway, e-mail: niklaska@oslomet.no"
date:         "2024-09-23"
bibliography: "../data/meta_data/references.bib"
csl:          "../scripts/university-of-gothenburg-apa-7th-edition-swedish-legislations.csl" # (source: https://bookdown.org/yihui/rmarkdown-cookbook/bibliography.html)
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true

@@

# cluster-variable
# kilde: https://www.r-bloggers.com/2021/05/clustered-standard-errors-with-r/
library(estimatr)                           # for cluster corrected SE
#library(sandwich)
#library(lmtest)
library(nlme)                               # for ICC - variance HLM

library(texreg)
```

@@

@@
